Countering language attrition with PanLex and the Web of Data

### 2014-02-18 panlex-rdf-paper/2012/MLOD_PanLex/paper_v4.tex
added paper_v4.tex

\author[A]{\fnms{Patrick} \snm{Westphal}},
\author[A]{\fnms{Claus} \snm{Stadler}},
\author[B]{\fnms{Jonathan} \snm{Pool}}
\address[A]{University of Leipzig, \{pwestphal, cstadler\}@informatik.uni-leipzig.de}
\address[B]{Long Now Foundation, San Francisco, pool@panlex.org}

\begin{abstract}
At present, there are approximately 7,000 living languages in the world.
However, some experts claim that the process of globalization may eventually lead to the world losing this linguistic diversity.
The vision of the PanLex project is to help save these languages, especially low-density ones, by allowing them to be intertranslatable and thus to be a part of the Information Age.
Semantic Web technologies can support achieving this goal, for reasons such as
their capabilities of flexibly representing, interlinking and reasoning with 
data, in our case particularly linguistic resources and annotations.
%Using Semantic Web technologies can support achieving this goal, especially due
% to inference capabilities and the interlinking of the PanLex data with other data sources.
Conversely, an RDF version of PanLex makes a significant contribution towards
improving the coverage of the Linguistic Web of Data, as to the best of our
knowledge there exists no large scale Linked Data data set for panlingual translation of
non-mainstream languages.
In this dataset description paper we detail how we transformed the data of the PanLex project to RDF,
established conformance with the lemon and GOLD data models,
interlinked it with Lexvo and DBpedia, and published it as Linked Data and via
SPARQL.
\end{abstract}

\begin{keyword}
Multi-lingual Linked Open Data, PanLex, Lexcial Resource, RDF, RDB2RDF, Sparqlify
\end{keyword}
\end{frontmatter}

\section{Introduction}
\label{sec:intro}
At present, there are about 7,000 living languages in the world\footnote{See \url{http://www.sil.org/iso639-3/download.asp} for a list of registered languages}.
Nonetheless, some experts claim that processes such as nation-state consolidation and globalization are producing language attrition so rapidly that up to 90\% of all languages alive today will be extinct within a century~\cite{endang_lang}.
Theorists of biolinguistic diversity argue that the loss of language diversity, the loss of human biological knowledge, and the loss of species diversity are mutually supportive and thus that language preservation and revitalization are essential to the preservation of biological diversity~\cite{nettle}.
The vision of the PanLex project is to help save these thousands of languages, especially those low-density ones that are threatened by extinction, by supporting their use in global communication.
This requires panlingual translation: translation from any language of the world into any other.
One of the crucial components of panlingual translation of discourses is panlingual lexical translation.
PanLex is designed to support that component by making use of several thousand
sources and artificial intelligence approaches.
%It documents the known lexical translations (translations of lexemes) among all
%languages.

Apart from using PanLex's data for inferring lexical translations, a
further direction of work is to generate added value by enriching this data with
knowledge from other (linguistic) sources.
Grounding on the idea of a Semantic Web, a complete stack of technologies has
been devised and standardized by the
\emph{World Wide Web Consortium}\footnote{\url{http://www.w3.org/standards/semanticweb/}}
supporting the definition of a machine readable and interpretable Linked Data
network, which has led to emergence of the \emph{Web of Data}.
%Technologies that support this aim are already available and in use in the so
%called \emph{Web of Data}.
Currently there is a growing community working on leveraging these Semantic Web
technologies for linguistic knowledge and thereby building a Web of Linguistic
Data, also known as the \emph{Linguistic Linked Open Data (LLOD) cloud}.

Our steps to connect PanLex to this Linked Data network are as follows.
In \autoref{sec:triplificate} we introduce the PanLex dataset, present our
PanLex RDF vocabulary, explain how we transformed the one into the other, and
how we established conformance with additional data models.
\autoref{sec:linking} is about how we linked to other datasets of the LLOD cloud,
whereas \autoref{sec:publishing} is about the actual dataset publishing.
Usage scenarios are given in \autoref{sec:usage}.
In \autoref{sec:related} we discuss related work, and finally, in \autoref{sec:conclusion} we conclude our approach and give some hints to future work.

\section{Triplification of the Raw Data}
\label{sec:triplificate}
In this section, we first provide an analysis of the PanLex dataset.
Subsequently, we introduce our URI and vocabulary design which closely
resembles PanLex's original conceptual model.
%, which was developed independently from other ontology engineering approaches.
Afterwards, we briefly describe how we classified PanLex's
instance data using additional data models. Finally, we explain the steps
taken to transform the data to RDF.

@@

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
At the core of the PanLex \emph{project}, there is the PanLex \emph{database} which is created from the imports of thousands of lexical resources, such as mono- and multilingual dictionaries, glossaries, standards, and thesauri.
The concrete list of used sources is available online\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.
The data derived from these sources comprise single- and multi-word expressions and meanings assigned to them.
Conceptually, the PanLex database thus ``represents \emph{assertions} about the \emph{meanings} of \emph{expressions}''\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
As of now, the database contains about 20 million meanings and 19 million expressions extracted from about 2,000 sources.
The most important entities and relations of PanLex's conceptual model are
depicted in \autoref{fig:db-schema} and are explained in more the detail in the following.

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{images/schema.png}
  \caption{The PanLex database schema}
  \label{fig:db-schema}
\end{figure}

\begin{itemize}
  \item The starting point of the data acquisition is the \emph{approver} entity: An editor processes the content of a certain mono- or multilingual source as mentioned above and adds it to the PanLex dataset.
    This combination of a user and a source is referred to as an approver.
  \item \emph{Expressions} are lemmas, i.e. dictionary entries.
    For example, \emph{``go''} is a valid expression, whereas \emph{``went''} is not.
    Expressions are always given in a language variety and can only be given once per language variety.
  \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
  \item \emph {Language varieties} allow one to make more fine grained distinctions within a language.
    Their codes are composed of the language code combined with a PanLex specific identifier.
    For example, ``eng'' is the ISO 639-3 code for the English language.
    Panlex defines various varieties of it, including ``English'' (eng-000), ``Simple English'' (eng-001) and ``British English'' (eng-005).
    Their labels, when possible, are autonyms, written in the native writing system.
    So, in contrast to mechanisms like IETF BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}} there is no need for a transcription.
  \item \emph{Meanings} in PanLex are entities of which each represents a unique possible sense of an expression.
    Meanings are assigned by editors based on their interpretation of expressions.
    Usually this assignment is done on a per-source basis so that identical meanings across multiple sources are not resolved.
    This means that if there is e.g. a translation of the fruit \emph{``apple''} in an English to German dictionary and another translation from English to French, these do not necessarily result in a single meaning entity linking to all involved languages.
    Instead, there could be \emph{``apple''} and \emph{``Apfel''} sharing one meaning entity and \emph{``apple''} and \emph{``pomme''} sharing another.
  \item \emph{Definitions} are optional descriptions of a meaning.
    They are given in a certain language variety.
    A description of the verb \emph{browse} for example could be \emph{``move or surf through various files on a computer, the Internet, etc.''}, marked as a definition in the ``English'' language variety.
  \item \emph{Denotations} are entities that relate expressions to meanings and may optionally carry annotations in form of sets of key value pairs.
    For instance, an English expression \emph{pig}, when referring to \emph{police officer}, could be annotated with \emph{pragmatics=vulgar}.
    Furthermore, denotations can be tagged with part-of-speech tags, such as word classes, selected from a closed list based on the Open Lexicon Interchange Format (OLIF) standard\footnote{\url{http://www.olif.net/}}.
    %For example, \emph{fall} can be a verb or a noun for autumn.
    %Homonyms are those expressions that are connected to multiple meanings.
  \item \emph{Users} have editorial privileges over the language varieties and the approvers that they define.
  \item \emph{Licenses} are also considered by the PanLex project.
    At present there are ten different license categories an approver can be annotated with.
    They are \emph{public domain}, \emph{Creative Commons (CC)}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License (GPL)}, \emph{GNU Lesser General Public License (LGPL)}, \emph{GNU Free Documentation License (FDL)}, \emph{MIT License}, \emph{copyright} (stating that there is a certain copyright holder), \emph{other} and \emph{unknown}.
    The license distribution is shown in \autoref{tbl:plx-license-counts}.
\end{itemize}

\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{lrclrclr}
License          & Count &&
License          & Count &&
License          & Count \\
\cline{1-2} \cline{4-5} \cline{7-8}
\emph{unknown}   &  1886 && % 1st
\emph{PD}        & 123 && % 2nd
\emph{LGPL}      &     9 \\ % 3rd
\emph{copyright} &  1212 && % 1st
\emph{other}     &   104 && % 2nd
\emph{request}   &     5 \\ % 3rd
\emph{CC}        &   335 && % 1st
\emph{MIT}       &    32 && % 2nd
\\
\emph{GPL}       &   172 && % 1st
\emph{FDL}       &    24 && % 2nd
\end{tabular}
\end{scriptsize}
\caption{Number of approvers using a certain license}
\label{tbl:plx-license-counts}
\end{table}

An overview of the number of instances per entity in the current PanLex database is given in~\autoref{fig:plx-entity-counts}.
\begin{table}
  \centering\begin{scriptsize}
  \begin{tabular}{lrclr}
    Entity             & Instances  &&
    Entity             & Instances  \\
    \cline{1-2} \cline{4-5}
    Denotations        & 50,803,243 &&  % SELECT COUNT(dn) FROM dn;
    Language Varieties &      7,248 \\  % SELECT COUNT(lv) FROM lv;
    Meanings           & 20,023,427 &&  % SELECT COUNT(mn) FROM mn;
    Approvers          &      3,905 \\  % SELECT COUNT(ex) FROM ex;
    Expressions        & 18,580,594 && % SELECT COUNT(ex) FROM ex;
    Licenses           &         10 \\ % SELECT COUNT(li) FROM apli;
    Definitions        &  2,522,605 && % SELECT COUNT(df) FROM df;
    Users              &          7 \\ % SELECT COUNT(us) FROM us;
    Languages          &      7,839 && % SELECT COUNT(lc) FROM lc; -- all language codes (ISO 639-3, ISO 639-2, ISO 639-5) known in the PanLex db
  \end{tabular}
  \end{scriptsize}
  \caption{Number of entities in the PanLex database}
  \label{fig:plx-entity-counts}
\end{table}

Note that \emph{approver} combines a user with an \emph{information source}, however the \emph{information source} is not modeled as a distinct entity.
Also, the information of whether or not two meanings with different approvers are the same is not being captured.

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
The entities and relations of the schema described in the previous section serve as the base for the development of the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside in the namespace \texttt{\small <http://ld.panlex.org/plx/>}, abbreviated with \texttt{\small plx}.
An example of the resulting ontology is depicted in \autoref{fig:vocabulary} and summarized as follows:
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern \emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case and the \{id\} is the primary key of the corresponding database table.

\begin{itemize}
  \item Expressions are modeled as instances of the class \texttt{\small plx:Expression}.
    Their original and normalized textual representations become the values of the properties \texttt{\small rdfs:label} and \texttt{\small plx:degradedText}, respectively.
    Their corresponding language variety is stated using \texttt{\small plx:languageVariety}.
  \item For language and language varieties the classes \texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
    \emph{ISO 639-1} and \emph{ISO 639-3 codes} become instances of the classes \texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code}.
  \item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
    Entities of this class may have an identifier assigned with the \texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
    Meanings may also have \emph{definitions}, entities of the \texttt{\small plx:Definition} class, giving a textual representation (\texttt{\small rdfs:label}) in a certain language variety (\texttt{\small plx:languageVariety}).
  \item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
    These are entities of the \texttt{\small plx:Denotation} class pointing to meanings and expressions via the properties \texttt{\small plx:denotationMeaning} and \\ \texttt{\small plx:denotationExpression}.
    Denotations may also have a word class assigned to them.
    This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
  \item All approvers share the \texttt{\small plx:Approver} class.
    The characteristics of an approver are described using mainly triples with literal objects.
    These are for example \texttt{\small dc:title} to assign the title of a source, \texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name.
\end{itemize}
Since the PanLex project compiled its database by extracting data from different sources, the licenses of these sources were also considered.
At present, we support the different license categories given in the database by creating resources of the \texttt{\small plx:License} class.

@@

\subsection{Transformation workflow}
\label{sec:conversion}
Since new sources are added to the PanLex database on almost a daily basis and because of its current size (\textasciitilde18 GB) an approach of recurrent conversions is impractical.
%Using conventional hardware, a full conversion takes infeasibly long.
%This makes the testing and debugging of the modeled vocabularies and data very
% time-consuming.
As the PanLex data already resides in a relational database, the use of a virtual RDB2RDF\footnote{\url{http://www.w3.org/2001/sw/wiki/RDB2RDF}} mapping solution is a natural choice.
The \emph{Sparqlify system}\footnote{\url{https://github.com/AKSW/Sparqlify}} offers, besides an efficient query rewriting engine, also a very easy-to-use mapping language, called \emph{Sparqlification Mapping Language} (SML).
Essentially, these mappings consist of three clauses:
The \emph{From} clause specifies the logical SQL table (i.e. table, view or
query) to be used in the SML view.
The \emph{With} clause binds a set of SPARQL variables to expressions that
yield RDF terms from relational columns.
Finally, the \emph{Construct} clause holds a set of triple patterns.
%clause making use of the defined SPARQL variable definitions.
\autoref{fig:ex:sparqlify-ml} shows an example of an SML view for the
languages in PanLex: From each row of the table \emph{i1} three resources are
created based on the \emph{iso3} column and 
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language} in the PanLex and the \emph{schema.org} namespace.
This view-based approach also demonstrates that changing the vocabulary or
adding support for new ones does not require an extract transform load (ETL) process,
and can therefore be done with little effort.

@@

\section{Related Work}
\label{sec:related}
PanLex is an integration project of many existing lexical resources.
The extraction of information from linguistic sources, and techniques for automatically inferring translations,
are relevant work discussed in \cite{panlex_probtrans}.
%A similar approach of linking expressions to meanings is followed by the
% \emph{Global Wordnet Association}.
An important initiative is the Global Wordnet Association\footnote{\url{http://www.globalwordnet.org/}\vspace{0.7em}},
which offers a platform for sharing wordnets and defines several goals. 
These include setting forth standards for uniformly representing wordnets
of different languages and establishing a universal index of meaning.
Wordnets are usually focused on the
definition of \emph{synsets} and relations between them in a single language,
whereas PanLex's focus is on capturing lexical translations between languages.
Hence, these efforts are complementary.
%As already mentioned in PanLex there can be several meaning entities
% representing the same unique sense.
In the Semantic Web context, several (quasi-)standard vocabularies and ontologies have been developed with the rise of the \emph{Linguistic Linked Open Data} movement.
Examples include the
\emph{Ontologies of Linguistic Annotation} (OLiA)~\cite{olia2010} for modeling
lexicon and machine-readable dictionaries, \emph{POWLA} for modeling linguistic corpora\cite{powla2012} and the \emph{Natural Language Processing Interchange Format} (NIF)\footnote{\url{http://nlp2rdf.org/nif-1-0}}.

\section{Conclusions and Future Work}
\label{sec:conclusion}
In this dataset description we detailed the PanLex database and its conversion to RDF.
Based on our URI and vocabulary design, we created appropriate view definitions
for the \emph{Sparqlify} system, which carries out the actual RDF
transformation.
Furthermore, we interlinked the languages in PanLex with Lexvo, and created about 2.5 million links to DBpedia for expressions in 16 languages.
With the integration of lemon and GOLD we also support the data access via external linguistic ontologies.

>>>>

\author[A]{\fnms{Patrick} \snm{Westphal}},
\author[A]{\fnms{Claus} \snm{Stadler}},
\author[B]{\fnms{Jonathan} \snm{Pool}}
\address[A]{University of Leipzig, \{pwestphal, cstadler\}@informatik.uni-leipzig.de}
\address[B]{The Long Now Foundation, San Francisco, pool@panlex.org}

%(%!TEX TS-program = xelatex)
%(%!TEX encoding = UTF-8 Unicode)

\begin{abstract}
At present, there are approximately 7,000 living languages in the world.
However, some forecasters expect the world to lose most of this linguistic diversity.
The mission of the PanLex project is to investigate whether it is feasible to make all languages intertranslatable and thus usable for world-wide communication, and, if so, whether that new situation will interfere with this expected language attrition.
This mission entails, among other prerequisites, making the lexicons of the world's languages intertranslatable. To accomplish that, the project is systematically integrating and making publicly available all known lexical translations.
Semantic Web technologies can enhance the power of PanLex by flexibly representing and reasoning with the content of its database and interlinking it with linguistic and other resources and annotations.
%Using Semantic Web technologies can support achieving this goal, especially due
% to inference capabilities and the interlinking of the PanLex data with other data sources.
Conversely, PanLex, with its unique collection of translation links between more than a billion pairs of lexemes from more than 9,000 language varieties, can improve the coverage of the Linguistic Web of Data.
In this dataset description paper we detail how we transformed the content of the PanLex database to RDF,
established conformance with the lemon and GOLD data models,
interlinked it with Lexvo and DBpedia, and published it as Linked Data and via
SPARQL.
\end{abstract}

\begin{keyword}
Multilingual Linked Open Data, LLOD Cloud, PanLex, Lexical Resource, RDF, RDB2RDF, SPARQL, Sparqlify
\end{keyword}
\end{frontmatter}

\section{Introduction}
\label{sec:intro}
At present, there are about 7,000 living languages in the world\footnote{\url{http://www-01.sil.org/iso639-3/iso-639-3.tab}. As of 25 January 2014, this list contained 6,975 living individual languages.}.
Nonetheless, processes such as urbanization, nation-state consolidation, and globalization appear to be producing language attrition, with, for example, the extinction or near-extinction in the last 60 years of from 10\% to over 75\% (depending on the world's region) of all living languages\cite{lang_crisis}.
Theorists of biolinguistic diversity argue that the loss of language diversity, the loss of human biological knowledge, and the loss of species diversity are mutually supportive and thus that language preservation and revitalization are essential to the preservation of biological diversity\cite{nettle}.

The mission of the PanLex project is to investigate the possibility that language attrition would be decreased or halted if utterances in any language were efficiently translatable into all other languages and, as a result, the speakers of all languages could use their languages for world-wide communication. Among the prerequisites of the panlingual translatablity of discourses would be the panlingual translatability of lexicons. In fact, lexical translation is almost the entirety of translation in some contexts (such as personal profiles, bibliographic subject headings, content tagging, product price lists, and web navigation controls). PanLex is designed to support that component by systematically integrating millions of lexical translations, documented in thousands of sources, into a database accessible to the research community, application developers, and the public. The database content can be interpreted as a graph linking millions of lexemes with one another in billions of ``is-a-translation-of'' relations, and permitting intelligent automated inference to trillions more such relations that have not been attested.
%It documents the known lexical translations (translations of lexemes) among all
%languages.

Precisely because the citation forms of lexemes, which is what PanLex directly enables the translation of, are commonly used as labels in collections of data, it is reasonable to speculate that PanLex might facilitate cross-linguistic access to knowledge. The idea of a Semantic Web has led to a stack of technologies standardized by the
\emph{World Wide Web Consortium}\footnote{\url{http://www.w3.org/standards/semanticweb/}}
supporting a machine-readable and machine-interpretable Linked Data
network. This has led to the concept and reality of a \emph{Web of Data}.
%Technologies that support this aim are already available and in use in the so
%called \emph{Web of Data}.
Currently there is a growing community working on leveraging these Semantic Web
technologies for linguistic knowledge and thereby building a Web of Linguistic
Data, also known as the \emph{Linguistic Linked Open Data (LLOD) cloud}.

Our steps to connect PanLex to this Linked Data network are as follows.
In \autoref{sec:triplificate} we introduce the PanLex dataset, present our
PanLex RDF vocabulary, and explain how we transformed the one into the other and established conformance with additional data models.
\autoref{sec:linking} is about how we linked to other datasets of the LLOD cloud,
and \autoref{sec:publishing} is about the publication of the dataset.
Usage scenarios are given in \autoref{sec:usage}.
In \autoref{sec:related} we discuss related work, and finally, in \autoref{sec:conclusion}, we conclude our approach and give some hints about future work.

\section{Triplification of the Raw Data}
\label{sec:triplificate}
In this section, we first provide an analysis of the PanLex dataset.
Subsequently, we introduce our URI and vocabulary design, which closely
resembles PanLex's original conceptual model.
%, which was developed independently from other ontology engineering approaches.
Afterwards, we briefly describe how we classified PanLex's
instance data using additional data models. Finally, we explain the steps
taken to transform the data to RDF.

@@

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
At the core of the PanLex \emph{project}, there is the PanLex \emph{database}, which is created by editors who consult thousands of lexical sources, such as mono- and multilingual dictionaries, glossaries, standards, and thesauri.
The list of sources that have been and are being consulted is available online\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.
The data based on these sources include single- and multi-word expressions, meanings assigned to them, and related information. Far from being mere copies of the sources, which would not make PanLex data interoperable, the data constitute editors' \emph{interpretations} of various sources' \emph{assertions} that two or more \emph{expressions} share a \emph{meaning}\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
As of January 2014, the database contained about 20 million meanings and 20 million expressions in about 9,200 language varieties based on the consultation of about 1,500 sources by about 20 editors.
The most important entities and relations of PanLex's conceptual model are depicted in \autoref{fig:db-schema} and are explained in more detail below.

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{images/schema-new.png}
  \caption{The PanLex database schema}
  \label{fig:db-schema}
\end{figure}

\begin{itemize}
  \item The starting point of data acquisition is the \emph{source} entity: An editor consults some mono- or multilingual source as mentioned above, interprets it as making assertions about lexical translations, and adds schema-compliant representations of these assertions to the PanLex dataset.
  \item \emph{Expressions} are lexical entities, each defined as a string of characters in some variety of some language. As a first approximation, expressions can be considered lemmas, i.e. dictionary-entry headwords. They differ from traditional headwords, however, in at least two ways. (1) Homographs, such as the verb ``hide'' (conceal) and the noun ``hide'' (animal skin) in English, are treated as a single expression in PanLex. (2) Multiword expressions, such as ``fall in love'', traditionally found in an entry headed by one of their words, such as ``fall'' or ``love'', are treated as independent expressions in PanLex. Expressions are strings of Unicode codepoints subjected to Normalization Form NFC, but this still makes it difficult to search precisely for expressions, so to assist in approximate searching PanLex also adds for each expression a degraded form permitting only lower-case letters and digits.
    \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes, and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
    \item \emph {Language varieties} are the fundamental linguistic entities in PanLex. Expressions are in them, not in languages. The language-variety concept is fine-grained and permissive: there can be many varieties of a language, and editors can add varieties as needed. A language variety is identified with some language code, combined with an arbitrary integer differentiating it from other varieties of the same language. For example, six dialects of Ahtna are identified as ``aht-000'' through ``aht-005''. These symbolic 3-letter, 3-digit labels are, themselves, treated as a (controlled) language variety, so editors can clarify their referents by translating them into natural-language expressions and into expressions in other controlled language varieties, such as the IETF standard BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}}. Such controlled language varieties may be defined as prohibiting synonymy and/or ambiguity.
    \item \emph{Meanings} in PanLex are entities with which expressions are semantically classified. An editor documents that two or more expressions are asserted to be translations or synonyms of one another by assigning a shared meaning to them. For example, depending on an editor's judgment, a source's translation of the German lexeme ``klingen'' into a sequence of alternative English lexemes ``ring, sound, seem'' can be documented by either (1) the assignment of all four expressions to the same meaning or (2) the assignment of two or three meanings to ``klingen'' and of one of those meanings to each of the English expressions. For practicality, PanLex treats expressions and meanings asymmetrically. Newly added translations automatically re-use existing matching expressions, but they don't re-use existing meanings, because there is no straightforward test of equivalence for meanings. PanLex generates new meanings as needed, and meanings are source-specific, leaving it for subsequent research to investigate which meanings should be considered equivalent and thus consolidated. Meanings can further have properties of three types. (1) \emph{Definitions} are descriptions of a meaning, consisting of arbitrary text strings annotated as being in particular language varieties. (2) \emph{Domains} are expressions (e.g., ``medicine'') that classify a meaning, but do not express it. (3) \emph{Meaning identifiers} are strings acting as references to identifiers used in a source.
    \item \emph{Denotations} are assignments of meanings to expressions. Each denotation, in addition to its meaning-expression pair, may be assigned one or more \emph{word classes} (a closed set based on OLIF, the Open Lexicon Interchange Format) and/or \emph{metadata} (arbitrary strings paired as keys and values).
    %For example, \emph{fall} can be a verb or a noun for autumn.
    %Homonyms are those expressions that are connected to multiple meanings.
  \item \emph{Users} are the entities that may act as editors by defining sources and then defining data attributed to those sources, as well as defining language varieties.
  \item Sources in PanLex are defined with about two dozen properties, among them being \emph{licenses}. As of January 2014 eleven different license categories were recognized. They were \emph{public domain}, \emph{Creative Commons (CC)}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License (GPL)}, \emph{GNU Lesser General Public License (LGPL)}, \emph{GNU Free Documentation License (FDL)}, \emph{MIT License}, \emph{copyright} (meaning that there is a copyright claim without a more liberal license also being offered), \emph{PanLex Use Permission} (where an author or publisher has given a source to PanLex for its use), \emph{other}, and \emph{unknown}.
    The distribution of licenses among sources that had been or were being consulted in January 2014 is shown in \autoref{tbl:plx-license-counts}.
\end{itemize}

\begin{table}
\centering
\begin{scriptsize}
\begin{tabular}{lrclrclr}
License          & Count &&
License          & Count &&
License          & Count \\
\cline{1-2} \cline{4-5} \cline{7-8}
\emph{copyright} &  1343  && \emph{LGPL}      &     9 && \emph{PD}           & 149 \\
\emph{CC}          &   387   && \emph{MIT}         &    32 && \emph{other}       &   106 \\
\emph{FDL}        &    24    && \emph{PanLex}    &    7 && \emph{unknown}  &  1958 \\
\emph{GPL}        &   172   && \emph{request}    &     5 \\
\end{tabular}
\end{scriptsize}
\caption{Number of sources using a certain license}
\label{tbl:plx-license-counts}
\end{table}

An overview of the number of instances per entity as of January 2014 is given in~\autoref{fig:plx-entity-counts}.
\begin{table}
  \centering\begin{scriptsize}
  \begin{tabular}{lrclr}
    Entity             & Instances  \\
    \cline{1-2}
    Denotations        & 54,278,860 \\  % SELECT COUNT(dn) FROM dn;
    Languages          &      7,843 \\ % SELECT COUNT(lc) FROM lc; -- all language codes (ISO 639-3, ISO 639-2, ISO 639-5) known in the PanLex db
    Language Varieties identified           &      9,310 \\  % SELECT COUNT(lv) FROM lv;
    Language Varieties with data           &      9,239 \\  % 
    Meanings           & 20,773,371 \\  % SELECT COUNT(mn) FROM mn;
    Sources being consulted         &      4,190 \\  % 
    Sources already consulted         &      1,453 \\  % 
    Expressions        & 19,790,453 \\ % SELECT COUNT(ex) FROM ex;
    Definitions        &  2,747,892 \\ % SELECT COUNT(df) FROM df;
    Users              &          23 \\ % SELECT COUNT(us) FROM us;
  \end{tabular}
  \end{scriptsize}
  \caption{Number of instances of main entities in the PanLex database}
  \label{fig:plx-entity-counts}
\end{table}

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
The entities and relations of the schema described in the previous section serve as the base for the development of the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside in the namespace \texttt{\small <http://ld.panlex.org/plx/>}, abbreviated with \texttt{\small plx}.
An example of the resulting ontology is depicted in \autoref{fig:vocabulary} and summarized as follows.
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern \emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case and the \{id\} is the primary key of the corresponding database table.

\begin{itemize}
  \item Expressions are modeled as instances of the class \texttt{\small plx:Expression}.
    Their original and degraded textual representations become the values of the properties \texttt{\small rdfs:label} and \texttt{\small plx:degradedText}, respectively.
    Their corresponding language variety is stated using \texttt{\small plx:languageVariety}.
  \item For language and language varieties the classes \texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
    \emph{ISO 639-1} and \emph{ISO 639-3 codes} become instances of the classes \texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code}.
  \item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
    Entities of this class may have an identifier assigned with the \texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
    Meanings may also have \emph{definitions}, entities of the \texttt{\small plx:Definition} class, giving a textual representation (\texttt{\small rdfs:label}) in a certain language variety (\texttt{\small plx:languageVariety}).
  \item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
    These are entities of the \texttt{\small plx:Denotation} class pointing to meanings and expressions via the properties \texttt{\small plx:denotationMeaning} and \\ \texttt{\small plx:denotationExpression}.
    Denotations may also have a word class assigned to them.
    This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
  \item All sources share the \texttt{\small plx:Approver} class.
    The characteristics of a source are described using mainly triples with literal objects.
    These are for example \texttt{\small dc:title} to assign the title of a source, \texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name. At present, we support the different license categories recognized in the database by creating resources of the \texttt{\small plx:License} class.
\end{itemize}

@@

\subsection{Transformation workflow}
\label{sec:conversion}
Since new sources are added to the PanLex database on almost a daily basis and because of its current size (\textasciitilde18 GB), the recurrent conversion of the database to capture changes in it is impractical.
%Using conventional hardware, a full conversion takes infeasibly long.
%This makes the testing and debugging of the modeled vocabularies and data very
% time-consuming.
As the PanLex data already reside in a relational database, the use of a virtual RDB2RDF\footnote{\url{http://www.w3.org/2001/sw/wiki/RDB2RDF}} mapping solution is a natural choice.
The \emph{Sparqlify system}\footnote{\url{https://github.com/AKSW/Sparqlify}} offers, besides an efficient query rewriting engine, also a very easy-to-use mapping language, called \emph{Sparqlification Mapping Language} (SML).
Essentially, these mappings consist of three clauses:
The \emph{From} clause specifies the logical SQL table (i.e. table, view, or
query) to be used in the SML view.
The \emph{With} clause binds a set of SPARQL variables to expressions that
yield RDF terms from relational columns.
Finally, the \emph{Construct} clause holds a set of triple patterns.
%clause making use of the defined SPARQL variable definitions.
\autoref{fig:ex:sparqlify-ml} shows an example of an SML view for the
languages in PanLex: From each row of the table \emph{i1} three resources are
created based on the \emph{iso3} column and 
bound to the variable names \emph{?lang}, \emph{?iso3} and \emph{?lexvo3}.
Resources for \emph{?lang} become typed as a \emph{Language} in the PanLex and the \emph{schema.org} namespace.
This view-based approach also demonstrates that changing the vocabulary or
adding support for new ones does not require an extract transform load (ETL) process,
and can therefore be done with little effort.

@@

\section{Related Work}
\label{sec:related}
PanLex is a project whose editors integrate information discovered from many lexical resources.
The extraction of information from linguistic sources, and techniques for automatically inferring translations,
are relevant work discussed in \cite{panlex_probtrans}.
%A similar approach of linking expressions to meanings is followed by the
% \emph{Global Wordnet Association}.
An important related initiative is the Global Wordnet Association\footnote{\url{http://www.globalwordnet.org/}\vspace{0.7em}},
which offers a platform for sharing wordnets and, among other goals, uniformly representing wordnets
of different languages and establishing a universal index of meaning.
Wordnets are usually focused on the
definition of \emph{synsets} and relations between them in a single language; GWA is helping to transform these single-language synonym silos into a virtual multilingual translation resource. PanLex is approximating that, in a different way, by integrating data from numerous wordnets along with translingual sources into a single graph.
%As already mentioned in PanLex there can be several meaning entities
% representing the same unique sense.
In the Semantic Web context, several standard or quasi-standard vocabularies and ontologies have been developed with the rise of the \emph{Linguistic Linked Open Data} movement.
Examples include the
\emph{Ontologies of Linguistic Annotation} (OLiA)~\cite{olia2010} for modeling
lexicon and machine-readable dictionaries, \emph{POWLA} for modeling linguistic corpora\cite{powla2012} and the \emph{Natural Language Processing Interchange Format} (NIF)\footnote{\url{http://nlp2rdf.org/nif-1-0}}.

\section{Conclusions and Future Work}
\label{sec:conclusion}
In this dataset description we detailed the PanLex database and its conversion to RDF.
Based on our URI and vocabulary design, we created appropriate view definitions
for the \emph{Sparqlify} system, which carries out the actual RDF
transformation.
Furthermore, we interlinked the languages in PanLex with Lexvo, and created about 2.5 million links to DBpedia for expressions in 16 languages.
With the integration of lemon and GOLD we also support data access via external linguistic ontologies.
